---
title: "A Brief Introduction to enpls"
author: "Nan Xiao <<https://nanx.me>>"
bibliography: enpls.bib
output:
  rmarkdown::html_document:
    toc: true
    toc_float: false
    toc_depth: 4
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A Brief Introduction to enpls}
---

# Introduction

The `enpls` package offers an algorithmic framework for measuring
feature importance, detecting outliers, and ensemble modeling
based on (sparse) partial least squares regression.
The key functions included in the package are listed in the table below.

+--------------------------------+-----------------------+------------------------------+
| Task                           | Partial Least Squares | Sparse Partial Least Squares |
+================================+=======================+==============================+
| Model fitting                  | `enpls.fit()`         | `enspls.fit()`               |
+--------------------------------+-----------------------+------------------------------+
| Cross validation               | `cv.enpls()`          | `cv.enspls()`                |
+--------------------------------+-----------------------+------------------------------+
| Detect outliers                | `enpls.od()`          | `enspls.od()`                |
+--------------------------------+-----------------------+------------------------------+
| Measure feature importance     | `enpls.fs()`          | `enspls.fs()`                |
+--------------------------------+-----------------------+------------------------------+
| Evaluate applicability domain  | `enpls.ad()`          | `enspls.ad()`                |
+--------------------------------+-----------------------+------------------------------+

Next, we will use the data from [@wang2015silico] to demonstrate
the general workflow of `enpls`. The dataset contains 1,000 compounds,
each characterized by 80 molecular descriptors. The response is the
octanol/water partition coefficient at pH 7.4 (logD7.4).

Let's load the data and take a look at it:

```{r}
library("enpls")
library("ggplot2")

data("logd1k")
x <- logd1k$x
y <- logd1k$y
head(x)[, 1:5]
head(y)
```

# Model Fitting

Here we fit the ensemble sparse partial least squares to the data,
so that the model complexity could usually be further reduced than
vanilla partial least squares when we build each model.

```{r, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
set.seed(42)
fit <- enspls.fit(x, y, ratio = 0.7, reptimes = 20, maxcomp = 3)
y.pred <- predict(fit, newx = x)

df <- data.frame(y, y.pred)
ggplot(df, aes_string(x = "y", y = "y.pred")) +
  geom_abline(slope = 1, intercept = 0, colour = "darkgrey") +
  geom_point(size = 3, shape = 1, alpha = 0.8) +
  coord_fixed(ratio = 1) +
  xlab("Observed Response") +
  ylab("Predicted Response")
```

We used the fitted model to predict on the training data and
plotted the predicted values against the true values.

The parameter `ratio` decides the sampling ratio for each Monte-Carlo
run; `maxcomp` controls the maximum number of components included
within each model; `reptimes` sets the times of Monte-Carlo resampling,
we recommend setting it to a large number (500 by default).

One common parameter for all functions in `enpls` is `parallel`,
it controls the number of CPU cores to use if you want to train
the models in parallel.

# Cross Validation

K-fold cross validation is a traditional way to measure the empirical
predictive performance of the model. We can use function `cv.enpls()`
or `cv.enspls()` to perform $k$-fold cross validation for the ensemble
(sparse) partial least squares model.

Since the parameters (number of components and level of sparsity)
are automatically tuned for each model in `enpls`, the cross validation
here is used to see if certain combinations of parameters (specified by
`ratio`, `maxcomp`, `alpha`, etc.) can produce ensemble models with
better performance.

```{r, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
cv.fit <- cv.enspls(x, y,
  nfolds = 5, ratio = 0.7,
  reptimes = 10, maxcomp = 3, verbose = FALSE
)
cv.fit
plot(cv.fit)
```

The returned object gives three model performance evaluation metrics
for the ensemble model: RMSE, MAE, and $R^2$. Here we also plotted the
predicted values for each test fold against the true response.

# Feature Importance

To measure feature importance, simply use `enpls.fs()` or `enspls.fs()`:

```{r, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
fs <- enspls.fs(x, y, ratio = 0.7, reptimes = 20, maxcomp = 3)
print(fs, nvar = 10)
plot(fs, nvar = 10)
plot(fs, type = "boxplot", nvar = 10)
```

The top 10 most important features are ranked as above. The boxplot
gives additional information about the coefficient stability of each
feature. We can see the feature TPSA (Topological Polar Surface Area)
has different pattern compared to others: it has large effect size,
but the effect sizes also have a large variance. This indicates
that TPSA is important for predicting logD7.4. However,
such importance may vary on different subsets of the samples.

# Outlier Detection

By using information from the prediction error distribution for each
sample produced by many models, we can measure if the responses
of particular samples are harder to predict than the others.
Such measurements can help on identifying outliers in the dataset.
Thus, they can be removed to get us a "clean" dataset before the
actual modeling.

This could be done with `enpls.od()`, `enspls.od()` easily:

```{r, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
od <- enspls.od(x, y, ratio = 0.8, reptimes = 20, maxcomp = 3)
plot(od, prob = 0.05)
plot(od, criterion = "sd", sdtimes = 2)
```

The two plots showed that several samples in our dataset might be
outlier candidates, based on two different criterions.
The samples in each area of the plots represent different types
of outliers.

# Applicability Domain Evaluation

Model applicability domain measures how well the predictive
model (PLS/Sparse PLS model) we built on the training set performs on
external test sets. A certain type of perturbation (such as bootstrapping,
jackknifing) is applied to the samples or the variables of the
training set, and we could get many different predictions for all
samples in all test sets (and of course, including the training set)
using each sub-model built with each perturbated training set.
The general evaluation strategy design and comparisons were analyzed
in @kaneko2014applicability.

The functions `enpls.ad()` and `enspls.ad()` could help us evaluate the
model applicability domain. Here we constructed two "pseudo" test sets
from the original `logd1k` dataset for demonstration:

```{r, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
# remove low variance variables
x <- x[, -c(17, 52, 59)]

# make training set
x.tr <- x[1:500, ]
y.tr <- y[1:500]

# make two test sets
x.te <- list(
  "test.1" = x[501:700, ],
  "test.2" = x[701:800, ]
)
y.te <- list(
  "test.1" = y[501:700],
  "test.2" = y[701:800]
)

ad <- enspls.ad(x.tr, y.tr, x.te, y.te,
  maxcomp = 3, space = "variable", method = "mc",
  ratio = 0.8, reptimes = 50
)
plot(ad)
```

Additionally, by using

```{r, eval = FALSE}
plot(ad, type = "interactive")
```

we will get an interactive plot, which could help us better explore
the model applicability domain by supporting zooming-in / inspecting
which sample each point represents interactively. The interactive plot
is based on [plotly](https://cran.r-project.org/package=plotly),
and it only requires an HTML viewer to be correctly rendered.

# Conclusion

Ensemble learning approaches are not only powerful for improving base
learner's predictive performance but also capable of accomplishing model
diagnostic tasks, such as measuring the importance of features.
It would be interesting to see if such ideas could be applied to more
relevant topics and further facilitate the predictive modeling tasks.

# References
